B\'{e}zier $\bar{B}$ projection

### 2018-01-24 projection1120.tex
Revisions.

\begin{abstract}
  We demonstrate the use of B\'{e}zier projection to alleviate locking phenomena in structural mechanics applications of isogeometric analysis. \added{Based on two different 	
  interpretations of well-known $\bar{B}$ projection, we propose two formulations and name them (symmetric/non-symmetric) B\'{e}zier $\bar{B}$ projection.} To demonstrate the utility of the \replaced{\Bezier projection}{approach} for both geometry and material locking phenomena we focus on transverse shear locking in Timoshenko beams and volumetric locking in nearly compressible linear elasticity although the approach can be applied generally to other types of locking phenemona as well. B\'{e}zier projection is a local projection technique with optimal approximation properties, which in many cases produces solutions that are comparable to global $L^2$ projection. In the context of $\bar{B}$ methods, the use of B\'ezier projection produces sparse stiffness matrices with only a slight increase in bandwidth when compared to standard displacement-based methods. Of particular importance is that the approach is applicable to any spline representation that can be written in B\'ezier form like NURBS, T-splines, LR-splines, etc. We discuss in detail how to integrate this approach into an existing finite element framework with minimal disruption through the use of B\'ezier extraction operators and a newly introduced dual \added{basis for} \Bezier \replaced{projection}{extraction} operator. We then demonstrate the behavior of the \replaced{two proposed formulations}{approach} through several challenging benchmark problems.
\end{abstract}
\keywords{Isogeometric analysis, B\'ezier extraction, \added{\Bezier dual basis}, B\'ezier projection, $\bar{B}$-projection, locking}
\section{Introduction}
Isogeometric analysis (IGA), introduced by Hughes et al. \cite{hughes_isogeometric_2005}, adopts the spline basis, which underlies the CAD geometry, as the basis for analysis. Of particular importance is the positive impact of smoothness on numerical solutions, where, in many application domains, IGA outperforms classical finite elements~\cite{cottrell_isogeometric_2009,cottrell_studies_2007,cottrell2006isogeometric,hughes_duality_2008,bazilevs_isogeometric_2010,evans_n-widths_2009}. Initial investigations of IGA focused on non-uniform rational B-splines (NURBS) due to their dominance in commercial CAD packages. However, many advances are being made in analysis-suitable geometry representations that overcome the strict rectangular topological restrictions of NURBS. Examples include T-splines~\cite{bazilevs_isogeometric_2010,sederberg_t-splines_2003} and their analysis-suitable restriction~\cite{scott_local_2012, li_analysis-suitable_2013}, hierarchical B-splines~\cite{bornemann_subdivision-based_2013,scott_isogeometric_2014,schillinger_isogeometric_2012,evans_hierarchical_2015,forsey_hierarchical_1988}, and locally refined B-splines~\cite{dokken_polynomial_2013,johannessen_isogeometric_2014} among others.

@@

∏

@@

and the term of this summation that results in the coupling between $N_{1}$ and $N_{5}$ is $\mathbf{P}_1^T\bar{\mathbf{M}}_2\mathbf{P}_3$, where $\mathbf{P}_1$ is the inner product of $N_{1}$ and $\hat{\bar{N}}_{2}$, $\mathbf{P}_3$ is the inner product of $N_{5}$ and $\hat{\bar{N}}_{3}$, and $\bar{\mathbf{M}}_2$ is the inner product of $\bar{N}_{2}$ and $\bar{N}_{3}$. We can see from Figure~\ref{fig:IGAelement} that $\operatorname{supp}(N_{1})\cap{}\operatorname{supp}(\hat{\bar{N}}_{2})=\Omega_1$, $\operatorname{supp}(N_{5})\cap{}\operatorname{supp}(\hat{\bar{N}}_{3})=\Omega_3$ and $\operatorname{supp}(\bar{N}_{2})\cap{}\operatorname{supp}(\bar{N}_{3})=\Omega_2$, so that $\mathbf{P}_1^T\bar{\mathbf{M}}_2\mathbf{P}_3$ is not zero. Thus we have increased the number of nonzero coefficients in the shear stiffness matrix. However, the same exercise can be used to show that there is no coupling between $N_{0}$ and $N_{6}$ for this set of basis functions so matrix is not dense. \added{With the abscense of the gramian matrix in the formulation of shear stiffness matrix, the bandwidth of the non-symmetric B\'ezier $\bar{B}$ method is even more reduced.} In fact, from the formulation of the element stiffness matrix, we can show that the bandwidth of the stiffness matrix of the symmetric B\'ezier $\bar{B}$ and non-symmetric B\'ezier $\bar{B}$ methods for the Timoshenko beam are $6p-3$ and $4p-1$, respectively.

@@

Following the same pattern of non-symmetric \Bezier $\bar{B}$ in Timoshenko beam problem, we keep the discretization of $\mathbf{u}$ and $\mathbf{w}$ untouched, while use lower order spline basis functions and corresponding dual basis functions for the discretization of pressure $p$ and its variation $\delta{p}$, respectively. The discretized stiffness matrix in mixed form can be written as
\begin{equation}
    \mathbf{K}_{mix}=
    \begin{bmatrix}
        \mathbf{K}^{dev} & -\mathbf{P}^{T}\\
        -\hat{\mathbf{P}} & -\dfrac{1}{(\lambda+2\mu/3)}\mathbf{I}
    \end{bmatrix},
\end{equation}
\added{where}
\begin{equation}
    \mathbf{P}=\A_e\mathbf{P}_e^{vol},
\end{equation}
with
\begin{equation}
    \mathbf{P}_e^{vol}=\bar{\mathbf{C}}^e\langle{\bar{\mathbf{B}}_e^T,\mathbf{B}^{vol}_e}\rangle(\mathbf{C}^e)^T.
\end{equation}
By eliminating the pressure control points, we obtain the pure displacement formulation as
\begin{equation}
	\mathbf{K} = \mathbf{K}^{dev} + \bar{\mathbf{K}}^{vol}_{NS},
\end{equation}
where
\begin{equation}
    \bar{\mathbf{K}}^{vol}_{NS}=(\lambda+2\mu/3){\mathbf{P}}^T\hat{\mathbf{P}}.
\end{equation}

\subsection{Numerical results}
We investigate the performance of the B\'{e}zier $\bar{B}$ method for two nearly incompressible linear elasticity problems under plane strain conditions. \added{Previous to the examples, we numerically evaluate the inf–sup constant for global $\bar{B}$ and non-symmetric \Bezier $\bar{B}$ methods. For numerical examples,} we first study the Cook's membrane problem, which is discretized with B-spline basis functions, and in the second problem we model the infinite plate with a circular hole problem using NURBS. Results computed using standard finite elements are labeled $Q_1$, $Q_2$, $Q_3$, and $Q_4$. Results computed using a global $\bar{B}$ method are labeled $\mathcal{T}^{L^2}$ and those computed with the \added{symmetric} B\'ezier $\bar{B}$ method and the \added{non-symmetric B\'ezier $\bar{B}$ method} are labeled $S-\mathcal{T}^{P}$ and $NS-\mathcal{T}^{P}$, respectively.

@@

Only the global $\bar{B}$ method and the non-symmetric \Bezier $\bar{B}$ method are considered here, as the symmetric \Bezier $\bar{B}$ method lacks connection to the mixed formulation. As a counter example, the well-known pair $Q_p/Q_p$ of global $\bar{B}$ that violates the inf-sup conditon is also tested here. Our tests follow the procedure proposed by Chapelle and Bathe in \cite{chapelle_inf-sup_1993}.\par
\begin{figure}[htb!]
    \centering
    \begin{subfigure}[b]{0.31\linewidth}        %% or \columnwidth
        \centering
        \includegraphics[width=\linewidth]{beta_p_2}
    \end{subfigure}
    \begin{subfigure}[b]{0.31\linewidth}        %% or \columnwidth
        \centering
        \includegraphics[width=\linewidth]{beta_p_3}
    \end{subfigure}
    \begin{subfigure}[b]{0.31\linewidth}        %% or \columnwidth
        \centering
        \includegraphics[width=\linewidth]{beta_p_4}
    \end{subfigure}

    \caption{Inf-sup test results for nearly incompressible elasticity. The stability parameter $\beta$ with respect to mesh size.}
    \label{fig:inf_sup}
\end{figure}

@@

\added{The error contour plots of $\sigma_{xx}$ for biquartic elements of the fifth mesh are shown in Figure \ref{fig:platewithhole_contour}, $Q_4$ elements are severely locked while all $\bar{B}$ methods converge to the reference solution. }
\begin{figure}[h]
    \center
\begin{tabular}{ccc}
\includegraphics[width=.31\linewidth]{deformation_p_2} & \includegraphics[width=.31\linewidth]{deformation_p_3} & \includegraphics[width=.31\linewidth]{deformation_p_4}\\
\includegraphics[width=.31\linewidth]{stress_p_2} & \includegraphics[width=.31\linewidth]{stress_p_3} & \includegraphics[width=.31\linewidth]{stress_p_4}\\
\includegraphics[width=.31\linewidth]{energy_p_2} & \includegraphics[width=.31\linewidth]{energy_p_3} & \includegraphics[width=.31\linewidth]{energy_p_4}
\end{tabular}
    \caption{Convergence study of the plate with a circular hole. The relative $L^2$ error of displacement, stress and the relative error in energy norm with respect to mesh refinement.}
	\label{fig:platewithhole_convergence}
\end{figure}

\begin{figure}[htb!]
    \center
    \begin{tabular}{cc}
    \includegraphics[width=.4\linewidth]{no} & \includegraphics[width=.4\linewidth]{g}\\
    $Q_4$ & $\mathcal{T}^{L^2} Q_4/Q_3$\\
    \includegraphics[width=.4\linewidth]{s} & \includegraphics[width=.4\linewidth]{ns}\\
    $S-\mathcal{T}^{p} Q_4/Q_3$ & $NS-\mathcal{T}^{p} Q_4/Q_3$
    \end{tabular}
        \caption{Contour plots of $\vert{\sigma_{xx}-\sigma_{xx}^h}\vert$ for the plate with a circular hole ($p=4$, the $5^{th}$ mesh is used).}
        \label{fig:platewithhole_contour}
    \end{figure}
\section{Conclusions}
\label{sec:conclusion}

>>>>

\begin{abstract}
  We demonstrate the use of B\'{e}zier projection to alleviate locking phenomena in structural mechanics applications of isogeometric analysis. \added{Based on two different 	
  interpretations of the well-known $\bar{B}$ projection, we propose two formulations and name them symmetric and non-symmetric B\'{e}zier $\bar{B}$ projection.} To demonstrate the utility of the \replaced{\Bezier projection}{approach} for both geometry and material locking phenomena we focus on transverse shear locking in Timoshenko beams and volumetric locking in nearly compressible linear elasticity although the approach can be applied generally to other types of locking phenemona as well. B\'{e}zier projection is a local projection technique with optimal approximation properties, which in many cases produces solutions that are comparable to global $L^2$ projection. In the context of $\bar{B}$ methods, the use of B\'ezier projection produces sparse stiffness matrices with only a slight increase in bandwidth when compared to standard displacement-based methods. Of particular importance is that the approach is applicable to any spline representation that can be written in B\'ezier form like NURBS, T-splines, LR-splines, etc. We discuss in detail how to integrate this approach into an existing finite element framework with minimal disruption through the use of B\'ezier extraction operators and a newly introduced dual \added{basis for} \Bezier \replaced{projection}{extraction} operator. We then demonstrate the behavior of the \replaced{two proposed formulations}{approach} through several challenging benchmark problems.
\end{abstract}
\keywords{Isogeometric analysis, B\'ezier extraction, \added{\Bezier dual basis}, B\'ezier projection, $\bar{B}$-projection, locking}
\section{Introduction}
Isogeometric analysis (IGA), introduced by Hughes et al. \cite{hughes_isogeometric_2005}, adopts the spline basis, which underlies the CAD geometry, as the basis for analysis. Of particular importance is the positive impact of smoothness on numerical solutions, where, in many application domains, IGA outperforms classical finite elements~\cite{cottrell_isogeometric_2009,cottrell_studies_2007,cottrell2006isogeometric,hughes_duality_2008,bazilevs_isogeometric_2010,evans_n-widths_2009}. Initial investigations of IGA focused on non-uniform rational B-splines (NURBS) due to their dominance in commercial CAD packages. However, many advances are being made in analysis-suitable geometry representations that overcome the strict rectangular topological restrictions of NURBS. Examples include T-splines~\cite{bazilevs_isogeometric_2010,sederberg_t-splines_2003} and their analysis-suitable restriction~\cite{scott_local_2012, li_analysis-suitable_2013}, hierarchical B-splines~\cite{bornemann_subdivision-based_2013,scott_isogeometric_2014,schillinger_isogeometric_2012,evans_hierarchical_2015,forsey_hierarchical_1988}, and locally refined B-splines~\cite{dokken_polynomial_2013,johannessen_isogeometric_2014} among others.

@@

âˆ

@@

and the term of this summation that results in the coupling between $N_{1}$ and $N_{5}$ is $\mathbf{P}_1^T\bar{\mathbf{M}}_2\mathbf{P}_3$, where $\mathbf{P}_1$ is the inner product of $N_{1}$ and $\hat{\bar{N}}_{2}$, $\mathbf{P}_3$ is the inner product of $N_{5}$ and $\hat{\bar{N}}_{3}$, and $\bar{\mathbf{M}}_2$ is the inner product of $\bar{N}_{2}$ and $\bar{N}_{3}$. We can see from Figure~\ref{fig:IGAelement} that $\operatorname{supp}(N_{1})\cap{}\operatorname{supp}(\hat{\bar{N}}_{2})=\Omega_1$, $\operatorname{supp}(N_{5})\cap{}\operatorname{supp}(\hat{\bar{N}}_{3})=\Omega_3$ and $\operatorname{supp}(\bar{N}_{2})\cap{}\operatorname{supp}(\bar{N}_{3})=\Omega_2$, so that $\mathbf{P}_1^T\bar{\mathbf{M}}_2\mathbf{P}_3$ is not zero. Thus we have increased the number of nonzero coefficients in the shear stiffness matrix. However, the same exercise can be used to show that there is no coupling between $N_{0}$ and $N_{6}$ for this set of basis functions so matrix is not dense. \added{With the abscense of the gramian matrix in the formulation of the shear stiffness matrix, the bandwidth of the non-symmetric B\'ezier $\bar{B}$ method is reduced further.} In fact, from the formulation of the element stiffness matrix, we can show that the bandwidth of the stiffness matrix of the symmetric B\'ezier $\bar{B}$ and non-symmetric B\'ezier $\bar{B}$ methods for the Timoshenko beam are $6p-3$ and $4p-1$, respectively.

@@

Following the same pattern as the non-symmetric \Bezier $\bar{B}$ method for the Timoshenko beam problem, we keep the discretization of $\mathbf{u}$ and $\mathbf{w}$ untouched, while using lower order spline basis functions and corresponding dual basis functions for the discretization of pressure $p$ and its variation $\delta{p}$, respectively. The discretized stiffness matrix in mixed form can be written as
\begin{equation}
    \mathbf{K}_{mix}=
    \begin{bmatrix}
        \mathbf{K}^{dev} & -\mathbf{P}^{T}\\
        -\hat{\mathbf{P}} & -\dfrac{1}{(\lambda+2\mu/3)}\mathbf{I}
    \end{bmatrix},
\end{equation}
\added{where}
\begin{equation}
    \mathbf{P}=\A_e\mathbf{P}_e^{vol},
\end{equation}
with
\begin{equation}
    \mathbf{P}_e^{vol}=\bar{\mathbf{C}}^e\langle{\bar{\mathbf{B}}_e^T,\mathbf{B}^{vol}_e}\rangle(\mathbf{C}^e)^T.
\end{equation}
By eliminating the pressure control points, we obtain the pure displacement formulation as
\begin{equation}
	\mathbf{K} = \mathbf{K}^{dev} + \bar{\mathbf{K}}^{vol}_{NS},
\end{equation}
where
\begin{equation}
    \bar{\mathbf{K}}^{vol}_{NS}=(\lambda+2\mu/3){\mathbf{P}}^T\hat{\mathbf{P}}.
\end{equation}

\subsection{Numerical results}
\added{We begin this section by numerically evaluating the inf€“-sup constant for the global $\bar{B}$ and non-symmetric \Bezier $\bar{B}$ methods.} 
We then investigate the performance of the B\'{e}zier $\bar{B}$ method for two nearly incompressible linear elasticity problems under plane strain conditions. \added{For the numerical examples,} we first study the Cook's membrane problem, which is discretized with B-spline basis functions, and in the second problem we model the infinite plate with a circular hole problem using NURBS. Results computed using standard finite elements are labeled $Q_1$, $Q_2$, $Q_3$, and $Q_4$. Results computed using a global $\bar{B}$ method are labeled $\mathcal{T}^{L^2}$ and those computed with the \added{symmetric} B\'ezier $\bar{B}$ method and the \added{non-symmetric B\'ezier $\bar{B}$ method} are labeled $S-\mathcal{T}^{P}$ and $NS-\mathcal{T}^{P}$, respectively.

@@

Only the global $\bar{B}$ method and the non-symmetric \Bezier $\bar{B}$ method are considered here, as the symmetric \Bezier $\bar{B}$ method lacks a connection to the mixed formulation. As a counter example, the well-known pair $Q_p/Q_p$ of the global $\bar{B}$ method that violates the inf-sup condition is also tested here. Our tests follow the procedure proposed by Chapelle and Bathe in \cite{chapelle_inf-sup_1993}.\par
\begin{figure}[htb!]
    \centering
    \begin{subfigure}[b]{0.31\linewidth}        %% or \columnwidth
        \centering
        \includegraphics[width=\linewidth]{beta_p_2}
    \end{subfigure}
    \begin{subfigure}[b]{0.31\linewidth}        %% or \columnwidth
        \centering
        \includegraphics[width=\linewidth]{beta_p_3}
    \end{subfigure}
    \begin{subfigure}[b]{0.31\linewidth}        %% or \columnwidth
        \centering
        \includegraphics[width=\linewidth]{beta_p_4}
    \end{subfigure}

    \caption{Inf-sup test results for nearly incompressible elasticity. The global $\bar{B}$ method, $\mathcal{T}^{L^2} Q4/Q3$, and the non-symmetric \Bezier $\bar{B}$ method, $NS - \mathcal{T}^P Q4/Q3$, do not strictly satisfy the LBB condition, but compared to the $\mathcal{T}^{L^2} Q_4/Q_4$ method, both methods reduce constraints to a favorable level. }
    \label{fig:inf_sup}
\end{figure}

@@

\added{Contour plots of the error in $\sigma_{xx}$ for biquartic elements of the most refined mesh are shown in Figure \ref{fig:platewithhole_contour}, We can see from these plots that $Q_4$ elements have severe error while the projection methods all reduce error to a more acceptable level. Here we can also see that the non-symmetric method has a slight improvement compared to the symmetric method.}
\begin{figure}[h]
    \center
\begin{tabular}{ccc}
\includegraphics[width=.31\linewidth]{deformation_p_2} & \includegraphics[width=.31\linewidth]{deformation_p_3} & \includegraphics[width=.31\linewidth]{deformation_p_4}\\
\includegraphics[width=.31\linewidth]{stress_p_2} & \includegraphics[width=.31\linewidth]{stress_p_3} & \includegraphics[width=.31\linewidth]{stress_p_4}\\
\includegraphics[width=.31\linewidth]{energy_p_2} & \includegraphics[width=.31\linewidth]{energy_p_3} & \includegraphics[width=.31\linewidth]{energy_p_4}
\end{tabular}
    \caption{Convergence study of the plate with a circular hole. The relative $L^2$ error of displacement, stress and the relative error in energy norm with respect to mesh refinement.}
	\label{fig:platewithhole_convergence}
\end{figure}

\begin{figure}[htb!]
    \center
    \begin{tabular}{cc}
    \includegraphics[width=.4\linewidth]{no} & \includegraphics[width=.4\linewidth]{g}\\
    $Q_4$ & $\mathcal{T}^{L^2} Q_4/Q_3$\\
    \includegraphics[width=.4\linewidth]{s} & \includegraphics[width=.4\linewidth]{ns}\\
    $S-\mathcal{T}^{p} Q_4/Q_3$ & $NS-\mathcal{T}^{p} Q_4/Q_3$
    \end{tabular}
        \caption{{\color{red} Can we change this to relative error?} Contour plots of $\vert{\sigma_{xx}-\sigma_{xx}^h}\vert$ for the plate with a circular hole ($p=4$, and the finest mesh is used).}
        \label{fig:platewithhole_contour}
    \end{figure}
\section{Conclusions}
\label{sec:conclusion}
